---
title: "Temperature dependence of consumer-resource interactions"
author: "Matt Barbour"
date: "`r Sys.Date()`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

```{r setup, include=FALSE}
## load libraries
library(deSolve) # numerical integration library
library(rootSolve) # for runsteady integration function
library(tidyverse) # for tidying data
library(cowplot) # for better base, ggplot graphics

# source in useful code
source("code/temperature-equations.R")
source("code/identify_steady_state.R")
source("code/CR_dynamics.R")

## set plot theme
theme_set(theme_cowplot())
```

## Consumer-resource model

I analyze a continuous time model of a consumer-resource interaction. The resource exhibits logistic growth and the consumer has a type 1 functional response:

$$\frac{dR}{dt}=R(r-\frac{rR}{K}-aC)$$
$$\frac{dC}{dt}=C(eaR-m)$$

```{r CR-model, include=FALSE}
CR_model <- function(Time,State,Pars) {
  with(as.list(c(State,Pars)), {
    # Consumer functional response
    CRfxn <- a * R 
    # population dynamics
    dR.dt <- r * R * (1 - R / K) - C * CRfxn
    dC.dt <- C * (e * CRfxn - m)
    # output  
    return(list(c(dR.dt, dC.dt)))
  })
}
```

The following ecological rates govern the dynamics of this model:

*r* = intrinsic growth rate of resource at low densities

*K* = carrying capacity of resource

*a* = attack rate of consumer on the resource

*e* = conversion efficiency of resources into consumers

*m* = mortality rate of consumer

I consider these ecological rates as the **phenotypes** of the consumer and resource.

## Scaling G$\times$E effects of temperature

Following the framework of @Gilbert2014, I assume that the temperature dependence of consumer and resource phenotypes can be modelled as Boltzmann-Arrhenius equations. I also use the biological plausible parameter values that they provide in Table 1 and Figure 3 of this paper. I explore the effects of temperature across a gradient from 5-30$^\circ$C.

```{r temp-gradient, include=F}
Temp_seq <- 5:30
```

To get a sense for how genotype-by-environment interactions (G$\times$E) might alter these dynamics, I created two different genotypes (A and B). These genotypes vary in their "initial" phenotype, which I setup at 15$^\circ$C. This choice of temperature was arbitrary and I chose it to follow Fig. 3 of @Gilbert2014. I would argue this represents the effect of genotype (G) as it affects the "intercept" of the phenotype. To simulate a G$\times$E effect, I adjusted the activation energy of the temperature relationship (i.e., the slope). With these two genotypes, I can visualize G effects (comparing at 15$^\circ$C), E effects (mean phenotype change with temperature), and G$\times$E effects (different slopes of each genotype).

Scaling intrinsic growth rate *r*:

```{r r-scaling}
# Genotype A
r_seq <- r_scaling(r0 = r0(r_base = 2, E_B = 0.32, T = C_to_K(15)), 
                   E_B = 0.32,
                   T = C_to_K(Temp_seq))

# Genotype B
r_seq2 <- r_scaling(r0 = r0(r_base = 1.5, E_B = 0.4, T = C_to_K(15)),
                   E_B = 0.4,
                   T = C_to_K(Temp_seq))
```

Scaling carrying capacity *K*:

```{r K-scaling}
# Genotype A
K_seq <- K_scaling(K0 = K0(K_base = 100, E_B = 0.32, E_S = 0.9, T = C_to_K(15)), 
                   E_B = 0.32,
                   E_S = 0.9,
                   T = C_to_K(Temp_seq))

# Genotype B
K_seq2 <- K_scaling(K0 = K0(K_base = 50, E_B = 0.4, E_S = 0.9, T = C_to_K(15)), 
                   E_B = 0.4,
                   E_S = 0.9,
                   T = C_to_K(Temp_seq))
```

Scaling mortality rate *m*:

```{r m-scaling}
# Genotype A
m_seq <- m_scaling(m0 = m0(m_base = 0.6, E_m = 0.45, T = C_to_K(15)), 
                   E_m = 0.45,
                   T = C_to_K(Temp_seq))

# Genotype B
m_seq2 <- m_scaling(m0 = m0(m_base = 0.65, E_m = 0.5, T = C_to_K(15)), 
                   E_m = 0.5,
                   T = C_to_K(Temp_seq))
```

Scaling of attack rate *a*:

```{r a-scaling}
# parameters from Gilbert et al. 2014 and Osmond et al. 2017

# Genotype A
a_seq <- a_scaling(a0 = a0(a_base = 0.1, v0_C = 1, v0_R = 1, 
                           E_vC = 0.46, E_vR = 0.46, 
                           T_C = C_to_K(15), T_R = C_to_K(15)),
                   v0_C = 1, v0_R = 1, E_vC = 0.46, E_vR = 0.46,
                   T_C = C_to_K(Temp_seq), T_R = C_to_K(Temp_seq))

# Genotype B
a_seq2 <- a_scaling(a0 = a0(a_base = 0.15, v0_C = 1, v0_R = 1, 
                            E_vC = 0.5, E_vR = 0.5, 
                            T_C = C_to_K(15), T_R = C_to_K(15)),
                   v0_C = 1, v0_R = 1, E_vC = 0.5, E_vR = 0.5,
                   T_C = C_to_K(Temp_seq), T_R = C_to_K(Temp_seq)) 
```

According to @Peters1983, conversion efficiency *e* is independent of temperature, so I set *e*=0.15 as in Fig. 3 of @Gilbert2014.

Below, I visualize the G$\times$E effects of temperature on each consumer and resource phenotype. This graph illustrates the well-known temperature dependence of each phenotype, but also indicates how we can model G$\times$E effects of temperature.

```{r plot-GxE, echo=FALSE, fig.height=2.5}
# organize phenotype data
r_df <- data.frame(Temp_seq, A = r_seq, B = r_seq2) %>%
  gather(key = Genotype, value = "phenotype", -Temp_seq) %>%
  mutate(parameter = "r")

K_df <- data.frame(Temp_seq, A = K_seq, B = K_seq2) %>%
  gather(key = Genotype, value = "phenotype", -Temp_seq) %>%
  mutate(parameter = "K")

m_df <- data.frame(Temp_seq, A = m_seq, B = m_seq2) %>%
  gather(key = Genotype, value = "phenotype", -Temp_seq) %>%
  mutate(parameter = "m")

a_df <- data.frame(Temp_seq, A = a_seq, B = a_seq2) %>%
  gather(key = Genotype, value = "phenotype", -Temp_seq) %>%
  mutate(parameter = "a")

e_df <- data.frame(Temp_seq, A = rep(0.15, length(Temp_seq)), B = rep(0.15, length(Temp_seq))) %>%
  gather(key = Genotype, value = "phenotype", -Temp_seq) %>%
  mutate(parameter = "e")

phenotype_df <- bind_rows(r_df, K_df, a_df, e_df, m_df) %>%
  mutate(Species = ifelse(parameter %in% c("r","K","a.R"), "resource", "consumer"))
phenotype_df$parameter <- factor(phenotype_df$parameter, levels = c("r","K","a","e","m"))

# plot phenotype data
plot_phenotype <- ggplot(phenotype_df, aes(x = Temp_seq, y = phenotype, color = Genotype, linetype = Species)) +
  geom_line() + 
  facet_wrap(~parameter, scales = "free_y", nrow = 1) +
#  geom_vline(xintercept = 15, linetype = "dotted", alpha = 0.5) +
  xlab("Temperature (C)") +
  ylab("Phenotype")
plot_phenotype
```

## How will G$\times$E effects of temperature alter natural selection?

To answer this question, I have to understand how a small change in the phenotype alters the mean fitness ($\bar{W}$) of the consumer or resource population (i.e. directional selection). Importantly, both consumer and resource fitness are density-dependent:

$$\bar{W_R}=\frac{1}{R}\frac{dR}{dt}=r-\frac{r}{K}R-aC$$
$$\bar{W_C}=\frac{1}{C}\frac{dR}{dt}=eaR-m$$

To make things easier, I assume the consumer and resource dynamics are at an equilibrium and use these values to to estimate selection on each phenotype.

```{r selection-dynamics, echo=FALSE}
# delta phenotype to estimate selection gradients 
delta_r <- 0.01
delta_K <- 0.01
delta_e <- 0.01
delta_m <- 0.01
delta_a <- 0.01

# selection on genotype A
selection_A <- CR_dynamics(
  init_parameters = data.frame(r = r_seq, K = K_seq, a = a_seq, e = 0.15, m = m_seq),
  init_states = data.frame(R = 1, C = 1),
  eco_CR_model = CR_model) %>%
  mutate(Genotype = "A",
         Temp_seq = Temp_seq,
         # calculate mean fitness for R
         R_base.fitness = r_seq - r_seq/K_seq*R - a_seq*C,
         # calculate mean fitness for C
         C_base.fitness = e*a_seq*R - m_seq,
         # calculate how a small change in parameter (delta_) influences mean fitness
         # note that the equilibrium R at the previous r value is assumed constant
         # this, I believe, is the same assumption for the Adaptive Dynamics approach,
         # but I'm unsure for quantitative genetics.
         
         # calculate selection gradient on r
         R_delta_r.fitness = (r_seq + delta_r) - (r_seq + delta_r)/K_seq*R - a_seq*C,
         beta_r = (R_delta_r.fitness - R_base.fitness)/delta_r,
         
         # calculate selection gradient on K
         R_delta_K.fitness = r_seq - r_seq/(K_seq + delta_K)*R - a_seq*C,
         beta_K = (R_delta_K.fitness - R_base.fitness)/delta_K,
         
         # calculate selection gradient on e
         C_delta_e.fitness = (e + delta_e)*a_seq*R - m_seq,
         beta_e = (C_delta_e.fitness - C_base.fitness)/delta_e,
         
         # calculate selection gradient on m
         C_delta_m.fitness = e*a_seq*R - (m_seq + delta_m),
         beta_m = (C_delta_m.fitness - C_base.fitness)/delta_m,
         
         # calculate selection gradient on a for Consumer
         C_delta_a.fitness = e*(a_seq + delta_a)*R - m_seq,
         beta_a.C = (C_delta_a.fitness - C_base.fitness)/delta_a,
         
         # calculate selection gradient on a for Resource
         R_delta_a.fitness = r_seq - r_seq/K_seq*R - (a_seq + delta_a)*C,
         beta_a.R = (R_delta_a.fitness - R_base.fitness)/delta_a)

# selection on genotype B
selection_B <- CR_dynamics(
  init_parameters = data.frame(r = r_seq2, K = K_seq2, a = a_seq2, e = 0.15, m = m_seq2),
  init_states = data.frame(R = 1, C = 1),
  eco_CR_model = CR_model) %>%
  mutate(Genotype = "B",
         Temp_seq = Temp_seq,
         # calculate mean fitness for R
         R_base.fitness = r_seq2 - r_seq2/K_seq2*R - a_seq2*C,
         # calculate mean fitness for C
         C_base.fitness = e*a_seq2*R - m_seq2,
         # calculate how a small change in parameter (delta_) influences mean fitness
         # note that the equilibrium R at the previous r value is assumed constant
         # this, I believe, is the same assumption for the Adaptive Dynamics approach,
         # but I'm unsure for quantitative genetics.
         
         # calculate selection gradient on r
         R_delta_r.fitness = (r_seq2 + delta_r) - (r_seq2 + delta_r)/K_seq2*R - a_seq2*C,
         beta_r = (R_delta_r.fitness - R_base.fitness)/delta_r,
         
         # calculate selection gradient on K
         R_delta_K.fitness = r_seq2 - r_seq2/(K_seq2 + delta_K)*R - a_seq2*C,
         beta_K = (R_delta_K.fitness - R_base.fitness)/delta_K,
         
         # calculate selection gradient on e
         C_delta_e.fitness = (e + delta_e)*a_seq2*R - m_seq2,
         beta_e = (C_delta_e.fitness - C_base.fitness)/delta_e,
         
         # calculate selection gradient on m
         C_delta_m.fitness = e*a_seq2*R - (m_seq2 + delta_m),
         beta_m = (C_delta_m.fitness - C_base.fitness)/delta_m,
         
         # calculate selection gradient on a
         C_delta_a.fitness = e*(a_seq2 + delta_a)*R - m_seq2,
         beta_a.C = (C_delta_a.fitness - C_base.fitness)/delta_a,
         
         # calculate selection gradient on a for Resource
         R_delta_a.fitness = r_seq2 - r_seq2/K_seq2*R - (a_seq2 + delta_a)*C,
         beta_a.R = (R_delta_a.fitness - R_base.fitness)/delta_a)
```

```{r plot-selection, include=FALSE}
# organize data
selection_df <- bind_rows(selection_A, selection_B) %>%
  select(Genotype, Temp_seq, beta_r, beta_K, beta_e, beta_m, beta_a.C, beta_a.R) %>%
  gather(key = "phenotype", value = "Beta", -Temp_seq, -Genotype) %>%
  separate(col = phenotype, into = c("beta","parameter"), sep = "_") %>%
  select(-beta) %>%
  mutate(Species = ifelse(parameter %in% c("r","K","a.R"), "resource", "consumer"),
         parameter = ifelse(parameter %in% c("a.R","a.C"), "a", parameter))
selection_df$parameter <- factor(selection_df$parameter, levels = c("r","K","a","e","m"))

plot_selection <- ggplot(selection_df, aes(x = Temp_seq, y = Beta, color = Genotype, linetype = Species)) +
  geom_line() +
  facet_wrap(~parameter, scales = "free_y", nrow = 1) +
  geom_hline(yintercept = 0) +
  #geom_vline(xintercept = 15, linetype = "dotted", alpha = 0.5) + 
  xlab("Temperature (C)") +
  ylab("Directional selection gradient")
```

Now I can plot both the G$\times$E effects of temperature, but also how these G$\times$E alters natural selection:

```{r plot-phenotype-selection, echo=FALSE}
plot_grid(plot_phenotype + xlab(""), plot_selection, nrow = 2)
```

This plot suggests several interesting patterns:

- Temperature increases *r* and *K* through a plastic response, but there is also positive directional selection on these traits. So in a sense, selection increase the magnitude of the environmental effect on *r* and *K*.

- Although temperature increases mortality rate exponentially, selection acts to decrease mortality rate in a way that is independent of temperature. Therefore, selection decreases the magnitude of the environmental effect on *m*.

- Despite conversion efficiency not changing with temperature, selection for greater conversion efficiency increases exponentially with temperature. This is indicative of temperature alter the ecological context, and thus selection on conversion efficiency.

Although it is not apparent, a close inspect of the above plot suggests that the absolute strength of selection is inversely related to the phenotype. To explore this explicitly, I plot this relationship below for the intialization temperature of 15$^\circ$C: 

```{r plot-selection-tradeoff, echo=FALSE}
phenotype_selection_df <- bind_rows(
  bind_rows(selection_A, selection_B) %>% select(Genotype, Temp_seq, phenotype = r, selection = beta_r) %>% mutate(parameter = "r"),
  bind_rows(selection_A, selection_B) %>% select(Genotype, Temp_seq, phenotype = K, selection = beta_K) %>% mutate(parameter = "K"),
  bind_rows(selection_A, selection_B) %>% select(Genotype, Temp_seq, phenotype = e, selection = beta_e) %>% mutate(parameter = "e"),
  bind_rows(selection_A, selection_B) %>% select(Genotype, Temp_seq, phenotype = a, selection = beta_a.C) %>% mutate(parameter = "a.C"),
  #bind_rows(selection_A, selection_B) %>% select(Genotype, Temp_seq, phenotype = a, selection = beta_a.R) %>% mutate(parameter = "a.R"),
  bind_rows(selection_A, selection_B) %>% select(Genotype, Temp_seq, phenotype = m, selection = beta_m) %>% mutate(parameter = "m")
) 

ggplot(filter(phenotype_selection_df, Temp_seq == 15), aes(x = log(phenotype), y = log(abs(selection)), color = Genotype)) + 
  geom_smooth(method = "lm", se = F) +
  geom_point(aes(shape = parameter)) +
  ggtitle("Temp = 15 C (initialization temperature)")
```

The above pattern is potentially interesting. Let's analyze it using a linear model:

```{r analyze-phenotype-selection}
summary(lm(log(abs(selection)) ~ Genotype + log(phenotype), 
           data = filter(phenotype_selection_df, Temp_seq == 15)))
```

The coefficient for log(phenotype) is ~1. Therefore, I believe this model implies that the strength of selection scales with the mean phenotype at a given temperature such that:

$$\beta\propto e^{-|z|}$$

I think this is interesting. It suggests that the contribution of selection to phenotypic change will be most important for phenotypes at smaller values.

How does this relationship change with temperature?

```{r analyze-phenotype-selection-temperature}
 # center at initializing T = 15 C
summary(lm(log(abs(selection)) ~ Genotype + I(Temp_seq - 15)*log(phenotype),
           data = phenotype_selection_df))
```

This model implies that that increasing temperature generally increases the strength of selection, but also alters the scaling relationship such that there is a greater contribution of smaller phenotypes, but weaker contribution of larger phenotypes (i.e. makes it more negative, see plot below).

```{r plot-phenotype-selection-temperature, echo=FALSE}
# I think the bad fits for Genotype B is weird. Looking at the stability graph below also makes me think
# that Genotype B is unusual.
ggplot(phenotype_selection_df, aes(x = log(phenotype), y = log(abs(selection)), group = Temp_seq)) + 
  geom_smooth(aes(color = Temp_seq), method = "lm", se = F) +
  geom_point(aes(shape = parameter)) +
  facet_wrap(~Genotype) +
  scale_color_viridis_c()
```

The departure of the lines for Genotype B at temperatures below 8$^\circ$C coincides with a funky stability pattern below. I don't quite understand this yet.

## Scaling G$\times$E effects to food-web stability

I can also look at how the G$\times$E effects of temperature alter the stability of the consumer-resource interactions:

```{r plot-stability, echo=FALSE}
# organize data
stability_df <- bind_rows(selection_A, selection_B) %>%
  select(Genotype, Temp_seq, max.Re.eigen) 

ggplot(stability_df, aes(x = Temp_seq, y = -1*max.Re.eigen, color = Genotype)) +
  geom_line() +
  geom_hline(yintercept = 0) +
  #geom_vline(xintercept = 15, linetype = "dotted", alpha = 0.5) + 
  xlab("Temperature (C)") +
  ylab("Stability (-1*max.Re.eigen)")
```

While we still observe a typical stability pattern with temperature [explained in detail in @Gilbert2014], we see that G$\times$E effects can switch which genotypes confer greater stability depending on temperature.

## Questions/Comments

- How do we incorporate the effects of selection onto the E effects of temperature? This visualization would be nice to evaluate the relative importance of selection vs E effects on the mean phenotype. Could it be as simple as adding the selection gradient multiplied by the additive genetic variance? e.g.

$$\Delta\bar{r}=r(T)+\beta_r\text{G}_r$$

- Although the above illustrates the effects of G$\times$E, it's still unclear to me whether selection favors Genotype A vs. B. Is this important to explore? Perhaps I could do this by using one of the genotypes as a reference (Genotype A) and then quantifying the relative fitness of Genotype B in the consumer-resource environment created by Genotype B (similar to an Adaptive Dynamics approach, where Genotype B would be the mutant invader).

- Is it worth plotting more genotypes? This could be done by simulating many initial parameter values for the G effects from a normal distribution, as well as the G$\times$E effect by simulating different activation energies for the temperature dependence of each phenotype.

- One thing that I do not address at all is the genetic covariance between these phenotypes (i.e. their G-matrix). This is certainly important for predicting evolutionary change in the next generation, but not critical for the selection analyses presented above. Interestingly, the selection analyses above could suggest what some of these genetic covariances might look like, since $\Delta\text{G}=\text{G}(\gamma-\beta\beta^\text{T})\text{G}$. Note that $\gamma$ refers to the matrix of nonlinear selection gradients, which I don't address here (perhaps I could assume they are negligible, since directional selection is strong? e.g. if directional selection is weak, then we might expect disruptive/stabilizing selection). Also T in this equation refers to the transpose of the vector of directional selection gradients ($\beta$) and not temperature.

## References

